Keep Your Laziness In Check

### 2017-11-19 slides.tex
backup slides
backup

\title{Keep Your Laziness In Check}
\author{\textbf{Kenny Foner} and \textbf{Hengchu Zhang}}
\institute{University of Pennsylvania}
\date{November 20, 2017}

@@

\begin{frame}
\frametitle{Being lazy is fun and useful, but\dots}
\Large\dots sometimes it leads to unintended consequences.
\end{frame}

\iffalse
\begin{frame}[fragile]
\frametitle{Amortized queues have strange laziness}
\begin{minted}{haskell}
enQ a (front,     back) =
      (front, a : back)

deQ ([],         [])   = Nothing
deQ (a : front', back) = Just (a, (front', back))
deQ ([],         back) = Just (a, (front', []))
  where
    (a : front') = reverse back
\end{minted}
\end{frame}

\note[itemize]{
  \item The usual implementation of functional queues uses two lists: when an
        item is enqueued, it is cons'ed onto the back list, and when we dequeue,
        an item is taken from the head of the front list. When the front list
        becomes empty, the back queue is reversed and takes the role of the new
        front list.
  \item This has amortized O(1) performance cost.
  \item This queue is lazy as long as you don't empty the front, then it is
        fully lazy in the structure of the back list. It only forces the spine
        of the back list when the front list is emptied.
}

\begin{frame}
\frametitle{But it doesn't have to be this way!}
\large
Chris Okasaki:\\
``Simple and Efficient Purely Functional Queues and Deques''
\vspace{.5\baselineskip}\\
(\emph{Journal of Functional Programming}, October 1995)
\end{frame}
\fi

\note[itemize]{
\item Enqueue O(1), Dequeue O(1) NOT AMORTIZED because of laziness!
\item Guarantees valid persistently; thunks are only evaluated once
\item How do we know we have implemented the Okasaki Queues correctly?
\item Correctness != functional correctness or micro-benchmark performance:
\item == the right amount of lazy
}

\begin{frame}
\frametitle{Wouldn't it be nice to QuickCheck laziness?}
Traditional property-based testing (such as QuickCheck):
\begin{itemize}
\item[\greentick] Great for testing functional correctness
\item[\greentick] Write a specification, fuzz inputs to functions to automatically test against that specification
\item[\redcross] Can't observe (or even specify) properties beyond functional correctness
\end{itemize}
If we were able to \textbf{specify} and \textbf{observe} laziness, we could
treat it \emph{just like} functional correctness.
\end{frame}

\note[itemize]{
\item QuickCheck gives functional correctness
\item But laziness can't be observed by QuickCheck
\item What if we make it observable? Then it's just part of functional correctness!
}

@@

\begin{frame}[fragile]
\frametitle{Observing strictness (part I)}
\begin{minted}{haskell}
instrumentListWithRef :: IORef Int -> [a] -> [a]
instrumentListWithRef _     []       = []
instrumentListWithRef count (a : as) =
  unsafePerformIO $ do
    modifyIORef' count (\x -> x + 1)
    return (a : instrumentListWithRef count as)
\end{minted}
\end{frame}

\note[itemize]{
\item \texttt{instrumentListWithRef} injected a snippet of code at every cons cell
\item and it returns a new ``loaded'' list
\item the injected code is executed as the ``loaded'' list gets pattern-matched on
}

\begin{frame}[fragile]
\frametitle{Strictness doesn't exist in a vacuum}

\begin{itemize}
\setlength\itemsep{-0.5em}
\item<1->[]
\begin{minted}{haskell}
type Context a = a -> ()
\end{minted}
\item<2->[]
\begin{minted}{haskell}
-----------------------------------------------
lazy, whnf, spineStrict :: Context [a]
lazy        = \xs -> const () xs
\end{minted}
\item<3->[]
\begin{minted}{haskell}
whnf        = \xs -> case xs of
                       []    -> ()
                       (_:_) -> ()
\end{minted}
\item<4->[]
\begin{minted}{haskell}
spineStrict =
  \xs -> whnf (foldl' (\ys y -> y : ys)) [] xs)
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
  \item The context type gives a function that returns a trivial value of the
        unit type
  \item It might look like there's only one such function \texttt{const ()}
  \item There can actually be all kinds of varying strictness behavior
  \item Forcing the unit value triggers the strictness behavior on the input \texttt{a}
  \item This gives us an easy way to observe the strictness behavior of a context
}

\begin{frame}[fragile]
\frametitle{Demanding an answer, lazily}
\begin{itemize}
\item<1->[]
\begin{minted}{haskell}
evaluate :: () -> IO ()
evaluate () = return ()
\end{minted}
\item<2->[]
\begin{minted}{haskell}
demandCount :: Context b -> ([a] -> b) -> [a] -> Int
demandCount c f as =
  unsafePerformIO $ do
    count <- newIORef 0
    let observableList =
          instrumentListWithRef count as
    evaluate ((c . f) observableList)
    readIORef count
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
  \item The seemingly trivial evaluate function forces the unit value returned
        by a context, thus triggering evaluation of the demanded part of
        observableList
  \item \texttt{demandCount} takes a context, and a function that operates over
        lists and the input list
  \item it applies instrumentListWithRef on the input list, producing an
        observableList, and applies the function and context over the
        observableList, triggering the injected instrumentation code
}

\begin{frame}[fragile]
\frametitle{Examples of \texttt{demandCount}}
\begin{itemize}
\item<1->[]
\begin{verbatim}
ghci> let f = take 6
\end{verbatim}
\item<2->[]
\begin{verbatim}
ghci> demandCount lazy f [1..]
0
\end{verbatim}
\item<3->[]
\begin{verbatim}
ghci> demandCount whnf f [1..]
1
\end{verbatim}
\item<4->[]
\begin{verbatim}
ghci> demandCount spineStrict f [1..]
6
\end{verbatim}
\end{itemize}
\end{frame}

\note[itemize]{
\item Just some examples showing demandCount
\item Note that we simply treat \texttt{f} as a black box
\item We don't need to know anything more about \texttt{f} other than the fact
      it typechecks with demandCount!
}

\begin{frame}[fragile]
\begin{columns}
\begin{column}{0.7\textwidth}
\frametitle{Beyond lists and numbers}
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{binary-tree}
\end{figure}
\end{column}
\begin{column}{0.2\textwidth}
{\fontsize{100}{50}\textrm{\textbf{?}}}
\end{column}
\end{columns}
\end{frame}

\note[itemize]{
\item This gives us a bare example that counts the number of cons cells forced,
      but what if we want more fine grained information of arbitrary algebraic
      data types?
\item Numbers can't directly capture the structure of a tree, so it can't
      capture the nodes of a tree that get forced in a computation
}

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand}}
\begin{minted}{haskell}
data List a =
    Cons a (List a)
  | Nil

data Thunk a = T | E a

data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data IntDemand = IntD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Examples}
\begin{minted}{haskell}
data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data IntDemand = IntD
---------------------------------------------------
ConsD T (ConsD (E IntD) (ConsD T T))

ConsD (E IntD) (ConstD T (ConsD (E IntD) (E NilD)))
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{\texttt{TreeDemand}}
\begin{minted}{haskell}
data Tree a =
    Node (Tree a) a (Tree a)
  | Leaf

data TreeDemand d =
    NodeD (Thunk (TreeDemand d))
          (Thunk d)
          (Thunk (TreeDemand d))
  | LeafD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand} vs \texttt{TreeDemand}}
\begin{minted}{haskell}
data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data TreeDemand d =
    NodeD (Thunk (TreeDemand d))
          (Thunk d)
          (Thunk (TreeDemand d))
  | LeafD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Computing demand, generically}
\begin{itemize}
\item<1->[]
\begin{minted}{haskell}
demandList :: Context b -> ([Int] -> b)
           -> [Int]
           -> (b, Thunk (ListDemand IntDemand))

demandTree :: Context b -> (Tree Int -> b)
           -> Tree Int
           -> (b, Thunk (TreeDemand IntDemand))
\end{minted}
\item<2->[]
\begin{minted}{haskell}
-----------------------------------------------
demand     :: Context b -> (a -> b) -> a
           -> (b, Thunk (Demand a))
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
\item FIX THE MARGIN ISSUES
\item There is a generic transformation between a data type and its
      corresponding data type representing a demand upon it
\item \texttt{Demand} is a type level function that calculates the demand type
\item We return a \texttt{Thunk} of demand because the input might not be
      demanded at all
\item As the input value is traversed by the function, which is driven by the
      context, the evaluation of each thunk builds a pointer based data
      structure which is isomorphic to the demand data type, which is then
      frozen into a demand representation
}

\begin{frame}[fragile]
\frametitle{Generic \texttt{Demand} calculation}
\begin{minted}{haskell}
type family Demand (x :: *) :: * where
 Demand (a -> b)  = FuncDemand
 Demand (a :+: b) = Demand a :+: Demand b
 Demand (a :*: b) = Thunk (Demand a) :*: Thunk (Demand b)
\end{minted}
\end{frame}

\note[itemize]{
\item For simplicity, we speak of types in terms of sums and products
\item \texttt{FuncDemand} is isomorphic to IntDemand in that it only captures
      whether the function was used at all
}

\begin{frame}[fragile]
\frametitle{First-order specifications}
\begin{minted}{haskell}
-- uncurried take
take     :: (Int, [a]) -> [a]

takeSpec :: Int -> [a]
         -> Demand [a]
         -> (Thunk (Demand Int), Thunk (Demand [a]))
\end{minted}
\end{frame}

@@

\note[itemize]{
  \item To QuickCheck \texttt{takeSpec}, we use mechanisms from QuickCheck to
        fuzz an Int and a [a] as inputs for \texttt{take}
  \item We use the \texttt{CoArbitrary} mechanism to generate random contexts
        that exerts non-trivial demand on the output values of \texttt{take}
  \item We use the generic \texttt{demand} function to kick off the entire
        computation
  \item We have now observed the demand on the input and also the demand on the
        output of take, so we can just straightforwardly compare that to what
        the specification expects.
  \item If we find a counterexample, we shrink the inputs first, and then shrink
        the reified demand, and re-exert that new demand on the output.
}

\begin{frame}[fragile]
\frametitle{Higher-order specifications}
\Huge PLACEHOLDER FOR PICTURE
\end{frame}

\note[itemize]{
  \item There is in fact also a mechanical transformation between the type of a
        function to the type of its demand specification
  \item Moreover, this transformation works even with high-order functions
  \item Higher-order specifications will take specification of their input
        functions because their demand behavior is parameterized over the demand
        behavior of their input functions
  \item At testing time, these specifications can be fuzzed through the same
        \texttt{CoArbitrary} mechanism
}

@@

>>>>

\title{Keep Your Laziness In Check}
\author{\textbf{Kenny Foner}, \textbf{Hengchu Zhang}, and \textbf{Leo Lampropoulos}}
\institute{University of Pennsylvania}
\date{November 20, 2017}

@@

\begin{frame}
\frametitle{Being lazy is sometimes vital, but\dots}
\Large\dots it can lead to unintended consequences.
\end{frame}

\note[itemize]{
\item Ubiquitous in Haskell programming, used in OCaml programming
\item (FRP) In the broader world, it is intrinsic in web-based functional
      reactive programming
\item (Spark) and in the field of big-data processing, for example, Spark
}

\begin{frame}[fragile]
\frametitle{Lazy lists}
\begin{minted}{ocaml}
type 'a lazyList =
  | Cons of ('a Lazy.t *
             'a lazyList Lazy.t)
  | Nil
\end{minted}
\end{frame}

\note[itemize]{
\item In this talk, we'll analyze functions that operate over lazy data structures
\item for instance, a lazy list like this
\item A lazy list is just a list with all of its parameters wrapped in a lazy
      computation
}

\begin{frame}[fragile]
\frametitle{Lazy queues}
\begin{minted}{ocaml}
let enQ a (front,                     back) =
          (front, lazy (Cons (lazy a, back)))

let deQ (front, back) =
  match Lazy.force front with
    | Cons (a, front') -> (a, (front', back))
    | Nil ->
        let Cons (a, front') = reverseLazyList back
        in (a, (front', lazy Nil))
\end{minted}
\end{frame}

\note[itemize]{
  \item As a micro example to motivate why thinking about laziness can be
        non-trivial, let's look at this queue
  \item In this slide, we use call-by-name ML for simplicity of presentation,
        but it can be implemented in OCaml using the Lazy module
  \item This has amortized O(1) performance cost.
  \item This queue is lazy as long as you don't empty the front, then it is
        fully lazy in the structure of the back list. It only forces the spine
        of the back list when the front list is emptied.
}

\iffalse
\begin{frame}
\frametitle{But it doesn't have to be this way!}
\large
Chris Okasaki:\\
``Simple and Efficient Purely Functional Queues and Deques''
\vspace{.5\baselineskip}\\
(\emph{Journal of Functional Programming}, October 1995)
\end{frame}

\note[itemize]{
\item Kenny:
\item Enqueue O(1), Dequeue O(1) NOT AMORTIZED because of laziness!
\item Guarantees valid persistently; thunks are only evaluated once
\item How do we know we have implemented the Okasaki Queues correctly?
\item Correctness != functional correctness or micro-benchmark performance:
\item == the right amount of lazy
}
\fi

\begin{frame}
\frametitle{Wouldn't it be nice to QuickCheck laziness?}
Traditional property-based testing (such as QuickCheck):
\begin{itemize}
\item[\greentick] Great for testing functional correctness
\item[\greentick] Write a specification, fuzz inputs to functions to automatically test against that specification
\item[\redcross] Can only specify and check functional correctness properties
\end{itemize}
If we were able to \textbf{specify} and \textbf{observe} laziness, we could
treat it \emph{just like} functional correctness.
\end{frame}

\note[itemize]{
\item QuickCheck gives functional correctness
\item But laziness can't be observed by QuickCheck
\item At the 3rd bullet point, say that laziness is not generally considered
      part of functional correctness much like how asymptotic runtime
      is not part of functional correctness because they are not directly
      observable
\item What if we make it observable? Then it's just part of functional correctness!
}

@@

\begin{frame}[fragile]
\frametitle{Observing strictness}
TODO: Just picture?
\begin{minted}{ocaml}
instrumentListWithRef : int ref
                      -> 'a lazyList
                      -> 'a lazyList
\end{minted}
\end{frame}

\note[itemize]{
\item Hengchu:
\item \texttt{instrumentListWithRef} clones the structure of the original list,
      and attaches a lazy effectful computation to each constructor in the list.
      So as the list is evaluated, each step of evaluation triggers one update
}

\begin{frame}[fragile]
\frametitle{Strictness doesn't exist in a vacuum}
TODO: Pictures with color to show how much is demanded
\end{frame}

\note[itemize]{
  \item These pictures are illustrations of a demand context on a lazy list
  \item These functions have the type from list to unit
  \item It might look like there's only one such function (namely the lazy
        implementation)
  \item There can actually be many different demand behaviors
  \item Forcing the unit value triggers the strictness behavior on the input \texttt{a}
  \item This gives us an easy way to observe the strictness behavior of a context
}

\begin{frame}[fragile]
\frametitle{Demanding an answer, lazily}
\begin{minted}{ocaml}
demandCount context f xs =
    let count          = ref 0
        observableList =
          instrumentListWithRef count xs
    in context (f observableList); !count
\end{minted}
\end{frame}

\note[itemize]{
  \item \texttt{demandCount} takes a context, and a function that operates over
        lists and the input list
  \item it applies instrumentListWithRef on the input list, producing an
        observableList, and applies the function and context over the
        observableList, triggering the injected instrumentation code
  \item context is exerting some demand on the output of f
}

\begin{frame}[fragile]
\frametitle{Examples of \texttt{demandCount}}
\begin{itemize}
\item<1->[]
\begin{verbatim}
> let f xs = takeLazy (lazy 6, xs)
> let list = toLazyList [1; 2; 3; 4; 5]
\end{verbatim}
\item<2->[]
\begin{verbatim}
> demandCount noDemand f list
\end{verbatim}
\item<3->[]
\begin{verbatim}
0
\end{verbatim}
\item<4->[]
\begin{verbatim}
> demandCount firstCons f list
\end{verbatim}
\item<5->[]
\begin{verbatim}
1
\end{verbatim}
\item<6->[]
\begin{verbatim}
> demandCount spineStrict f list
\end{verbatim}
\item<7->[]
\begin{verbatim}
5
\end{verbatim}
\end{itemize}
\end{frame}

\note[itemize]{
\item Just some examples showing demandCount
\item takeLazy is a function that takes the specified number of elements from a
      lazy list and returns a lazyList
\item Note that we simply treat \texttt{f} as a black box
\item We don't need to know anything more about \texttt{f} other than the fact
      it typechecks with demandCount!
}

\begin{frame}[fragile]
TODO: erase the numbers in tree
\begin{columns}
\begin{column}{0.7\textwidth}
\frametitle{Beyond lists and numbers}
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{binary-tree}
\end{figure}
\end{column}
\begin{column}{0.2\textwidth}
{\fontsize{100}{50}\textrm{\textbf{?}}}
\end{column}
\end{columns}
\end{frame}

\note[itemize]{
\item Kenny:
\item Arbitrary lazy algebraic datatypes
\item This gives us a bare example that counts the number of cons cells forced,
      but what if we want more fine grained information of arbitrary algebraic
      data types?
\item Numbers can't directly capture the structure of a tree, so it can't
      capture the nodes of a tree that get forced in a computation
}

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand}}
\begin{minted}{ocaml}
type 'a lazyList =
  | Cons of ('a Lazy.t *
             'a lazyList Lazy.t)
  | Nil
------------------------------------
type 'a thunk = T | E of 'a

type 'd listDemand =
  | ConsD of ('d thunk *
              'd listDemand thunk)
  | NilD

type intDemand = IntD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Examples}
\begin{minted}{ocaml}
type 'd listDemand =
  | ConsD of ('d thunk *
              'd listDemand thunk)
  | NilD

type intDemand = IntD
---------------------------------------------------
ConsD (T,
  ConsD (E IntD,
    ConsD (T, T)))

ConsD (E IntD,
  ConsD (T,
    ConsD (E IntD, E NilD)))
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{\texttt{TreeDemand}}
\begin{minted}{ocaml}
type 'a lazyTree =
  | Node of ('a lazyTree Lazy.t *
             'a Lazy.t *
             'a lazyTree Lazy.t)
  | Leaf

type 'd treeDemand =
  | NodeD of ('d treeDemand thunk *
              'd thunk *
              'd treeDemand thunk)
  | LeafD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand} vs \texttt{TreeDemand}}
\begin{minted}{ocaml}
type 'd listDemand =
  | ConsD of ('d thunk *
              'd listDemand thunk)
  | NilD

type 'd treeDemand =
  | NodeD of ('d treeDemand thunk *
              'd thunk *
              'd treeDemand thunk)
  | LeafD
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Computing demand, generically}
\begin{itemize}
\item<1->[]
\begin{minted}{ocaml}
demandList : 'b context -> (int lazyList -> 'b)
           -> int lazyList
           -> ('b * intDemand listDemand thunk)

demandTree : 'b context -> (int lazyTree -> 'b)
           -> int lazyTree
           -> ('b * intDemand treeDemand thunk)
\end{minted}
\item<2->[]
\begin{minted}{ocaml}
------------------------------------------------
demand     : 'b context -> ('a -> 'b) -> 'a
           -> ('b * ('a DEMAND) thunk)
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
\item There is a determinate relation between a lazy data structure and its
      demand representation
\item We return a \texttt{Thunk} of demand because the input might not be
      demanded at all
\item As the input value is traversed by the function, which is driven by the
      context, the evaluation of each thunk builds a pointer based data
      structure which is isomorphic to the demand data type, which is then
      frozen into a demand representation
\item We expect that the 'a parameter is some kind of lazy structure, and 'b is
      also some kind of lazy structure
\item We use generic programming to implement the DEMAND type for all algebraic
      data types
}

\iffalse
\begin{frame}[fragile]
\frametitle{Generic \texttt{Demand} calculation}
\begin{minted}{haskell}
type family Demand (x :: *) :: * where
 Demand (a -> b)  = FuncDemand
 Demand (a :+: b) = Demand a :+: Demand b
 Demand (a :*: b) = Thunk (Demand a) :*: Thunk (Demand b)
\end{minted}
\end{frame}

\note[itemize]{
\item For simplicity, we speak of types in terms of sums and products
\item \texttt{FuncDemand} is isomorphic to IntDemand in that it only captures
      whether the function was used at all
}
\fi

\begin{frame}[fragile]
\frametitle{First-order specifications}
\begin{minted}{ocaml}
deQSpec :  (int lazyList Lazy.t * int lazyList Lazy.t)
        -> (intDemand thunk *
             (intDemand listDemand thunk *
              intDemand listDemand thunk))
        -> (intDemand listDemand thunk *
            intDemand listDemand thunk)
deQSpec (front, back) (dInt, (dFront, dBack)) =
  match Lazy.force front with
    | Cons (_, _) -> (E (ConsD (dInt, dFront)), dBack)
    | Nil -> (dBack,
              (spineStrictD |back|)
               U ((spineStrictD
                   (|back| - |dFront|))
                  @ (reverse (ConsD dInt dFront)))
\end{minted}
\end{frame}

@@

\note[itemize]{
  \item To QuickCheck \texttt{takeSpec}, we use mechanisms from QuickCheck to
        fuzz an Int and a [a] as inputs
  \item We can generate random contexts
        that exerts non-trivial demand on the output values
  \item We use the generic \texttt{demand} function to kick off the entire
        computation
  \item We have now observed the demand on the input and also the demand on the
        output of take, so we can just straightforwardly compare that to what
        the specification expects.
  \item If we find a counterexample, we shrink the inputs first, and then shrink
        the reified demand, and re-exert that new demand on the output.
  \item Higher-order also works! but we don't want to go into the details
}

\iffalse
\begin{frame}[fragile]
\frametitle{Higher-order specifications}
\Huge PLACEHOLDER FOR PICTURE
\end{frame}

\note[itemize]{
  \item There is in fact also a mechanical transformation between the type of a
        function to the type of its demand specification
  \item Moreover, this transformation works even with high-order functions
  \item Higher-order specifications will take specification of their input
        functions because their demand behavior is parameterized over the demand
        behavior of their input functions
  \item At testing time, these specifications can be fuzzed through the same
        \texttt{CoArbitrary} mechanism
}
\fi

@@

\begin{frame}[fragile]
\frametitle{Have questions?}
Come ask us about \dots
\begin{itemize}
\item Higher-order specifications
\item Generic programming
\item Random generation of demand contexts
\item Details of instrumentation
\item What language are you actually implementing this in? \\ (Hint: it's not ML)
\item[] \dots
\end{itemize}
\end{frame}
